---
output:
  html_document: default
  pdf_document: default
  title: "Data Partitioning"
  date: '`r paste("First created on June 01, 2023. Updated on", Sys.Date())`'
---

```{r echo = FALSE}
knitr::opts_knit$set(root.dir = dirname(dirname(rstudioapi::getSourceEditorContext()$path)))
```

Read in SparkR functions from **create_partitions.R**.
```{r}
library(dplyr)
library(devtools)
source('walkingcalculatr/R/create_partitions.R')
```

## Spark Overview

**Apache Spark** is an open-source, distributed processing system used for big data workloads. **Sparklyr**[^1] is an R package that provides bindings to Spark’s distributed machine learning library.  Sparklyr is normally preferred over **SparkR**[^2] (Spark's wrapper library for R) as it includes dplyr compatibility however, for data partitioning which uses read/write operations SparkR has faster performance. 

## Spark Installation

SparkR is included with a version of Spark installed. A version of Spark can be installed manually via the [Apache Spark website](https://spark.apache.org/downloads.html) or using the `spark_install()` function from the sparklyr package.

### Installing Spark from Sparklyr

sparklyr can be installed from CRAN as follows. 
```{r, eval = FALSE}
install.packages('sparklyr', INSTALL_opts = '--no-lock', quiet = F)
```

sparklyr makes it easy to install spark using R. 
A list of available versions can be printed.
```{r}
library(sparklyr)
print(spark_available_versions())
```

Download a recent version of Spark using `spark_install()`.
```{r, eval = FALSE}
spark_install(version = "3.3.0")
```

### Java Installation

SparkR runs on both Windows and UNIX-like systems (e.g. Linux, Mac OS), and it should run on any platform that runs a supported version of Java (version between 8 - 17). Java can be manually installed from this [link](https://www.oracle.com/java/technologies/downloads/#java8). 

Once Java installed, the JAVA_HOME environment variable should be updated with the java installed path.

For MacOS, this can be done via Terminal:
```
export JAVA_HOME="path/to/your/java/jdk"
```

For Windows, this can be done via Command Prompt:
```
setx JAVA_HOME "path/to/your/java/jdk" /M
```

A user can also set the JAVA_HOME variable on their R session prior to initialing a Spark session:
```{r, eval = FALSE}
Sys.setenv(JAVA_HOME = "path/to/your/java/jdk")
```

## SparkR Setup

To find the directory a version of Spark is installed, create a Spark connection (aka Session) 
using sparklyr and create a variable with the **spark_home** dir. If the Spark directory is known, this step is optional.
```{r, warning= F}
library(sparklyr)

sc <- spark_connect(master = "local")
spark_path = sc$spark_home
spark_disconnect(sc)
```

Given a version of Spark installed, the SparkR library can be imported as follows. Replace **spark_path** with your Spark home directory and create a temporary directory to avoid memory issues
```{r, echo = F, warning = F}
# Read in SparkR package given spark home dir
Sys.setenv(SPARK_HOME = spark_path)

# Set spark local directories to avoid memory issues on /tmp
Sys.setenv(R_SESSION_TMPDIR = './spark-tmp')
library(SparkR, lib.loc = c(file.path(Sys.getenv("SPARK_HOME"), "R", "lib")))
```

A Spark session can be initialized with multiple configuration options such as 
`spark.driver.memory` which allocates memory for the current workers. More information on available Spark configuration parameters can be found [here](https://spark.apache.org/docs/latest/sparkr.html#starting-up-sparksession).

```{r, echo = F, warning = F}
sc <- sparkR.session(master = "local[*]",
                     appName = "LBS_R",
                     sparkConfig = c(spark.driver.memory = "16g",
                                     spark.local.dir = '/home/balderama/Rscratch/spark-tmp',
                                     spark.sql.session.timeZone = "UTC",
                                     spark.sql.broadcastTimeout = "3600"))

sq <- sparkRSQL.init(sc)
```

## SparkR Usage


To read in data as a spark data frame, the `read.df()` function can be used.
A helper function `read_csvs()` which casts all csvs within a directory (**\csvPath**) to a spark data frame was created for your convenience. See the example below:
```{r,  eval = F}
# Helper function to read in raw csvs and concatenate
read_csvs <- function(file_lst) {
  
  # Read in first file
  df <- read.df(file_lst[[1]], "csv",
                header = "true",
                inferSchema = "true",
                na.strings = "NA")
  
  # Concatenate with remaining
  for (i in seq_along(file_lst)) {
    
    if (i > 1) {
      temp <- read.df(file_lst[[i]],
                      "csv",
                      header = "true",
                      inferSchema = "true",
                      na.strings = "NA")
      
      df <- rbind(df, temp)
    }
  }
  return(df)
}

# Read in data
df_raw <- read_csvs(csvPath)
```

For our example, we will read in synthetic LBS data from our package as follows:
```{r,  eval = F}
devtools::load_all(".")
data("synthetic_ness_data")
```

Then cast the data to a spark data frame.
```{r,  eval = F}
df_raw <- as.DataFrame(synthetic_ness_data)
```

The data schema can be printed out for a given spark data frame.
```{r,  eval = F}
# Show schema
printSchema(df_raw)
```

Basic operations such as displaying and obtaining counts can also be performed.
```{r,  eval = F}
# Show counts
SparkR::count(df_raw)
```

```{r,  eval = F}
# Display data
head(df_raw)
```

In addition, data wrangling using a syntax similar to dplyr can be used. 
```{r,  eval = F}
# Lower case all character vectors to ensure de-duping
df_raw$id <- SparkR::lower(df_raw$id)

# Down-select and rename some rows
df_clean <- df_raw %>% SparkR::select('id', 
               'unix_timestamp',
               'latitude', 
               'longitude') %>% 
                SparkR::withColumnRenamed("id", "userId") %>% 
                SparkR::withColumnRenamed("unix_timestamp", "time") %>% 
                SparkR::withColumnRenamed("latitude", "lat") %>% 
                SparkR::withColumnRenamed("longitude", "lon") 

# Remove duplicates by id and time
df_clean <- df_clean %>% SparkR::dropDuplicates(c('userId', 'time'))

# Show updated counts
SparkR::count(df_clean)
```

SQL queries can be run on your spark data frame as well. 
SQL queries execute faster then native SparkR functions, so is the preferred method for basic calculations.
```{r,  eval = F}
# Cast data-frame as temp view
createOrReplaceTempView(df_clean, "df_clean")

# Get number of records
numDevices <- SparkR::collect(SparkR::sql("SELECT COUNT(*) as TOT_ROWS from df_clean"))

# Get number of unique devices
numUnique <- SparkR::collect(SparkR::sql("SELECT COUNT(DISTINCT(userId)) from df_clean"))

# Get average, max and min num of pings by device
numPings <- SparkR::collect(SparkR::sql("SELECT MEAN(NumPings), MAX(NumPings), MIN(NumPings) FROM (SELECT userId, COUNT(*) as NumPings from df_clean GROUP BY userId)"))

# Get counts by bbox
numLats <- SparkR::collect(SparkR::sql("SELECT COUNT(*) FROM from df_clean WHERE lat BETWEEN 80 AND 81"))

# Print out summary table
summary_stats <- cbind(numDevices, numUnique, numPings)
head(summary_stats)
```

Data can be written to local memory using `write.parquet()` or `write.csv()` functions.
```{r,  eval = F}
# Save as parquet with snappy compression (to save space)
write.parquet(df_clean, "../data/spark_ness_parquet", 
              mode = "overwrite", 
              compression = "snappy")
```

Confirm data written to local memory can be read in.
```{r, eval = F}
df <- SparkR::read.df(sc, 
                      path = "../data/spark_ness_parquet",
                      source = 'parquet', 
                      schema = NULL)

printSchema(df)
head(df)
```

A Spark session should be disconnected once spark is no longer needed as follows. 
```{r, eval = F}
# Disconnect Spark session
sparkR.stop()
```

## Partitioning Data

Partitioning is only necessary if the current size of the data cannot be read in memory using conventional R packages (data.table, dplyr). Currently the walkingcalculatr functions have been verified to work on LBS data  (< 200 million rows) given a  64 GB Linux machine. As read/write operations have slower performance, the smallest number of N partitions is recommended. A helper script **create_partitions.R** is included to help with partitioning the data.

The `create_partition_from_parquet()` function ensures N partitions from a given LBS directory are created
with device stratified partitions. Device stratification is required for walkingcalculatr functions as 
all records within a given device must be located within the same file. 

```{r,  eval = F}
# Create N partitions and save to output dir
create_partition_from_parquet(sc, 
                              sq,
                              "../data/spark_ness_parquet",
                              "../data/spark_ness_partitions",
                              N = 10)
```
